\bf An Examination of Sport Climbing Competition Format and Scoring System

### 2021-07-29 manuscript/Rmd/draft.Rmd
Update Rmd

---
title: An Examination of Sport Climbing Scoring System

@@

keywords:
- sport climbing
- scoring system

abstract: |
  The purpose of this paper is to investigate sport climbing, a new sport featured in the 2020 Tokyo Summer Olympics. Simulation. Data Analysis: drop and re-rank, correlation
# bibliography: bib_example.bib
output: rticles::asa_article
---

\normalsize

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = "",
                      # fig.width = 5,
                      # fig.height = 4,
                      results = "asis")
options(xtable.comment = FALSE,
        xtable.caption.placement = "top", 
        xtable.include.rownames = FALSE)
```

# Introduction (sport climbing, other scoring system, mention rank product)

@@

In the Olympic sport climbing, there are 20 competitors at the start (in both men's and women's).  All 20 competitors compete in each of the three events in the qualification round, and their performances in each event are ranked from 1 to 20.  A competitors final score is then computed as the product of their ranks in the three events and the lower product is better.  Specifically, $$Score_i = R^S_i\times R^B_i\times R^L_i,$$ where $R^S_i$, $R^B_i$, and $R^L_i$ are the ranks of the $i$-th competitor in speed climbing, bouldering, and lead climbing, respectively.

@@

@@

## Data Collection

We collected data on major climbing competitions from 2018 to 2020, including the 2020 Continental Championships of Europe, Africa, Oceania, Pan-America; 2019 and 2018 World Championships; 2018 Asian Games; and 2018 Asian Games.

## EDA

@@

We conducted a simulation study to examine the rankings and scoring for climbers in both qualification and final rounds. For each round, we performed 10000 simulations, and this was accomplished by randomly assigning the ranks of each event to every participant, with the assumption that the ranks are uniformly distributed. After the completion of the simulations, we calculated the final scores for every simulated round, as well as the final standings for the climbing athletes. This data would then enable us to answer questions about various topics, including the distributions of scores for qualifying and final rounds, and the probabilities of advancing to the finals or winning a medal, given certain conditions.

```{r, fig.cap = "Distribution of qualification scores"}
qual %>%
  ggplot(aes(qual_score)) +
  geom_histogram(bins = 20, color = "white") + 
  labs(x = "Qualification Score",
       y = "Frequency")
  # theme(axis.title = element_text(size = 9),
  #       axis.text = element_text(size = 8))
```

For the qualification round, a climber is almost guaranteed to make the final round if they win the first event (with a 99.51% chance of advancing) or if they win at least one of the three climbing concentrations (99.48%). On the other hand, finishing last in the first event or in any event would certainly hurt an athlete's chance of finishing in the top 8, as the probabilities of a climber advancing given they finish last in the first and in any event are 0.1830 and 0.1885, respectively. In addition, the average score for qualification positions 1 to 8 are displayed in Table 1. We notice that on average, the minimum score that one should aim for in order to move on to the final round is 435 (for 8th rank).

@@

```{r, fig.cap = "Kendall's rank correlations - 2018 World Championship, Women's Qualification"}
wq %>% 
  select(rank, speed, bould, lead) %>% 
  ggpairs(diag = "blank",
          upper = list(continuous = wrap("cor", method = "kendall")))
```

@@

```{r}
women_final <- read_csv("https://raw.githubusercontent.com/qntkhvn/climbing/main/data/2018_youth_olympics/women_final.csv")
rerank <- list()
for (i in 1:nrow(women_final)){
  rerank[[i]] <- women_final[-i,] %>%
    mutate(rank_drop = i)
}

rerank_df <- women_final %>% 
  mutate(rank_drop = 0) %>% 
  bind_rows(rerank) %>% 
  group_by(rank_drop) %>% 
  mutate(speed = rank(speed),
         bould = rank(bould),
         lead = rank(lead),
         total = speed * bould * lead) %>% 
  arrange(total, .by_group = TRUE) %>% 
  ungroup() %>% 
  group_by(rank_drop, total) %>% 
  mutate(speed_tb = ifelse(speed < lag(speed), 1, 0),
         bould_tb = ifelse(bould < lag(bould), 1, 0),
         lead_tb = ifelse(lead < lag(lead), 1, 0),
         tb = speed_tb + bould_tb + lead_tb,
         tb = ifelse(is.na(tb), 1, tb)) %>% 
  ungroup() %>% 
  group_by(rank_drop) %>% 
  arrange(total, -tb, .by_group = TRUE) %>% 
  mutate(rank = row_number())
```

```{r, fig.cap = "Original rankings and rankings after each rank is dropped"}
rerank_df %>% 
  mutate(last = fct_reorder(last, -rank)) %>% 
  # filter(rank_drop %in% c(0, 5)) %>% 
  ggplot(aes(x = last, y = total)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = rank), hjust = -0.2, size = 3) +
  coord_flip() +
  facet_wrap(~ rank_drop) +
  expand_limits(y = 65) +
  labs(y = "Score",
       x = "Climber")
```

>>>>

---
title: An Examination of Sport Climbing Competition Format and Scoring System

@@

keywords:
- sport climbing
- scoring system
- 2020 Summer Olympics

abstract: |
  The purpose of this paper is to investigate sport climbing, one of the sports making its debut on the Olympics stage at Tokyo 2020. In particular, we take a closer look at the controversial competition format and scoring system of sport climbing.

  ... Simulation. Data Analysis: drop and re-rank, correlations
# bibliography: bib_example.bib
output: rticles::asa_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = "",
                      fig.height = 4,
                      fig.width = 5,
                      results = "asis")
options(xtable.comment = FALSE,
        xtable.caption.placement = "top", 
        xtable.include.rownames = FALSE)
```

# Introduction

@@

In the Olympic sport climbing, there are 20 competitors at the start (in both men's and women's).  All 20 competitors compete in each of the three events in the qualification round, and their performances in each event are ranked from 1 to 20.  A competitors final score is then computed as the product of their ranks in the three events and the lower product is better.  Specifically, 
\begin{equation}
Score_i = R^S_i\times R^B_i\times R^L_i,
\end{equation} 
where $R^S_i$, $R^B_i$, and $R^L_i$ are the ranks of the $i$-th competitor in speed climbing, bouldering, and lead climbing, respectively.

@@

As a side note, rank product, genes

@@

We collected data on major climbing competitions from 2018 to 2020, including the 2020 Continental Championships of Europe, Africa, Oceania, Pan-America; 2019 and 2018 World Championships; 2018 Asian Games; and 2018 Youth Olympics. Data were obtained from various sources, including the event websites, Wikipedia, and the International Federation of Sport Climbing (IFSC). The main attributes of our datasets are the name and nationality of the climbers; bib number (for some competitions); the finishing place of climbers in speed climbing, bouldering, and lead climbing; the total score (which equals the product of event ranks); and the final rank. We utilize this data to compute the correlations between the event ranks and final table position, as well as to look at how often the final orderings change if one athlete is dropped and the ranks for each discipline are re-computed.

EDA

Corr

Drop 1

@@

We conducted a simulation study to examine the rankings and scoring for climbers in both qualification and final rounds. For each round, we performed 10000 simulations, and this was accomplished by randomly assigning the ranks of each event to every participant, with the assumption that the ranks are uniformly distributed. After the completion of the simulations, we calculated the total scores for every simulated round, as well as the final standings for the climbing athletes. Table 
and
Figure
are numerical and visual summaries of final scores obtaining from our simulated data for the qualification round.

```{r}
mosaic::favstats(qual$qual_score)[-9] %>% 
  xtable(caption = "Descriptive statistics of simulated qualification scores") %>% 
  print()
```

```{r, fig.cap = "Histogram of qualification scores"}
qual %>%
  ggplot(aes(qual_score)) +
  geom_histogram(bins = 20, color = "white") + 
  labs(x = "Qualification Score",
       y = "Frequency")
  # theme(axis.title = element_text(size = 9),
  #       axis.text = element_text(size = 8))
```

The simulation results allow to answer questions about various topics, including the distributions of scores for qualifying and final rounds, and the probabilities of advancing to the finals or winning a medal, given certain conditions. For the qualification round, a climber is almost guaranteed to make the final round if they win the first event (with a 99.51% chance of advancing) or if they win at least one of the three climbing concentrations (99.48%). On the other hand, finishing last in the first event or in any event would certainly hurt an athlete's chance of finishing in the top 8, as the probabilities of a climber advancing given they finish last in the first and in any event are 0.1830 and 0.1885, respectively. In addition, the average score for qualification positions 1 to 8 are displayed in Table 1. We notice that on average, the minimum score that one should aim for in order to move on to the final round is 435 (for 8th rank).

@@

```{r, fig.cap = "Kendall's rank correlations - 2018 World Championship, Women's Qualification", fig.width = 6}
wq %>% 
  select(rank, speed, bould, lead) %>% 
  ggpairs(diag = "blank",
          axisLabels = "internal",
          upper = list(continuous = wrap("cor", method = "kendall", stars = FALSE))) +
  theme(panel.grid.major = element_blank())
```

@@

```{r}
wf <- read_csv("https://raw.githubusercontent.com/qntkhvn/climbing/main/data/2018_youth_olympics/women_final.csv")
mf <- read_csv("https://raw.githubusercontent.com/qntkhvn/climbing/main/data/2018_youth_olympics/men_final.csv")

drop_rerank <- function(df) {
  rerank <- list()
  for (i in 1:nrow(df)) {
    rerank[[i]] <- df[-i, ] %>%
      mutate(rank_drop = i)
  }

  rerank_df <- df %>%
    mutate(rank_drop = 0) %>%
    bind_rows(rerank) %>%
    group_by(rank_drop) %>%
    mutate(
      speed = rank(speed),
      bould = rank(bould),
      lead = rank(lead),
      total = speed * bould * lead
    ) %>%
    arrange(total, .by_group = TRUE) %>%
    ungroup() %>%
    group_by(rank_drop, total) %>%
    mutate(
      speed_tb = ifelse(speed < lag(speed), 1, 0),
      bould_tb = ifelse(bould < lag(bould), 1, 0),
      lead_tb = ifelse(lead < lag(lead), 1, 0),
      tb = speed_tb + bould_tb + lead_tb,
      tb = ifelse(is.na(tb), 1, tb)
    ) %>%
    ungroup() %>%
    group_by(rank_drop) %>%
    arrange(total,-tb, .by_group = TRUE) %>%
    mutate(rank = row_number(),
           last = str_to_title(last))

  return(rerank_df)
}
```

```{r, fig.cap = "This figure shows the rankings of the Men's Sport Climbing Final at the 2018 Youth Olympics. Each panel represents the dropped rank, with 0 being the original final standings. The cases where a change in ordering happened are highlighted.", fig.width = 7}
drop_rerank(mf) %>% 
  mutate(last = fct_reorder(last, -rank),
         rank = as.factor(rank),
         rank_change = ifelse(
           rank_drop %in% c(0, 4) | 
             rank_drop == 1 & rank %in% 4:5 |
              rank_drop == 2 & rank %in% c(1, 5) |
                rank_drop %in% c(3, 5) & rank %in% c(1, 2, 5) |
                  rank_drop == 6 & rank %in% 1:3, 
           "no", "yes")) %>% 
  ggplot(aes(x = last, y = total, fill = rank_change)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = rank), hjust = -0.2, size = 3) +
  coord_flip() +
  geom_rect(aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf),
            data = ~ filter(., rank_drop %in% c(1, 2, 3, 5, 6)), 
            color = "black", size = 1.5, fill = NA, inherit.aes = FALSE) +
  facet_wrap(~ rank_drop, nrow = 2) +
  expand_limits(y = 102, x = 0:7) +
  scale_fill_manual(values = c("grey", "chocolate")) +
  labs(y = "Score",
       x = "Climber") +
  theme_minimal() +
  theme(axis.ticks = element_blank(),
        panel.grid.minor = element_blank())
```

```{r, fig.cap = "This figure shows the final rankings of 2018 Youth Olympics Women's Final. Each panel represents the dropped rank, with 0 being the original final standings. The cases where a change in ordering happened are highlighted.", fig.width = 7}
drop_rerank(wf) %>% 
  mutate(last = fct_reorder(last, -rank),
         rank = as.factor(rank),
         rank_change = ifelse(
           rank_drop %in% c(0, 1, 4, 6) |
             rank_drop == 2 & rank %in% 1:3 |
              rank_drop == 3 & rank %in% c(1, 4, 5) |
                rank_drop == 5 & rank %in% c(1, 5),
           "no", "yes")) %>%
  ggplot(aes(x = last, y = total, fill = rank_change)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = rank), hjust = -0.2, size = 3) +
  coord_flip() +
  geom_rect(aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf),
            data = ~ filter(., rank_drop %in% c(2, 3, 5)), 
            color = "black", size = 1.5, fill = NA, inherit.aes = FALSE) +
  facet_wrap(~ rank_drop, nrow = 2, ncol = 4) +
  expand_limits(y = 62, x = 0:7) +
  scale_fill_manual(values = c("grey", "chocolate")) +
  labs(y = "Score",
       x = "Climber") +
  theme_minimal() +
  theme(axis.ticks = element_blank(),
        panel.grid.minor = element_blank())
```

### 2021-07-29 manuscript/Rmd/draft.tex
Update Rmd

% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{textcomp}
\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

\if0\blind
{
  \title{\bf An Examination of Sport Climbing Scoring System}

@@

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf An Examination of Sport Climbing Scoring System}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
The purpose of this paper is to investigate sport climbing, a new sport
featured in the 2020 Tokyo Summer Olympics. Simulation. Data Analysis:
drop and re-rank, correlation
\end{abstract}

\noindent%
{\it Keywords:} sport climbing, scoring system
\vfill

@@

\normalsize

\hypertarget{introduction-sport-climbing-other-scoring-system-mention-rank-product}{%
\section{Introduction (sport climbing, other scoring system, mention
rank
product)}\label{introduction-sport-climbing-other-scoring-system-mention-rank-product}}

@@

In the Olympic sport climbing, there are 20 competitors at the start (in
both men's and women's). All 20 competitors compete in each of the three
events in the qualification round, and their performances in each event
are ranked from 1 to 20. A competitors final score is then computed as
the product of their ranks in the three events and the lower product is
better. Specifically, \[Score_i = R^S_i\times R^B_i\times R^L_i,\] where
\(R^S_i\), \(R^B_i\), and \(R^L_i\) are the ranks of the \(i\)-th
competitor in speed climbing, bouldering, and lead climbing,
respectively.

@@

\hypertarget{data-collection}{%
\subsection{Data Collection}\label{data-collection}}

We collected data on major climbing competitions from 2018 to 2020,
including the 2020 Continental Championships of Europe, Africa, Oceania,
Pan-America; 2019 and 2018 World Championships; 2018 Asian Games; and
2018 Asian Games.

\hypertarget{eda}{%
\subsection{EDA}\label{eda}}

@@

We conducted a simulation study to examine the rankings and scoring for
climbers in both qualification and final rounds. For each round, we
performed 10000 simulations, and this was accomplished by randomly
assigning the ranks of each event to every participant, with the
assumption that the ranks are uniformly distributed. After the
completion of the simulations, we calculated the final scores for every
simulated round, as well as the final standings for the climbing
athletes. This data would then enable us to answer questions about
various topics, including the distributions of scores for qualifying and
final rounds, and the probabilities of advancing to the finals or
winning a medal, given certain conditions.

\begin{figure}
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{Distribution of qualification scores}
\end{figure}

For the qualification round, a climber is almost guaranteed to make the
final round if they win the first event (with a 99.51\% chance of
advancing) or if they win at least one of the three climbing
concentrations (99.48\%). On the other hand, finishing last in the first
event or in any event would certainly hurt an athlete's chance of
finishing in the top 8, as the probabilities of a climber advancing
given they finish last in the first and in any event are 0.1830 and
0.1885, respectively. In addition, the average score for qualification
positions 1 to 8 are displayed in Table 1. We notice that on average,
the minimum score that one should aim for in order to move on to the
final round is 435 (for 8th rank).

@@

\begin{figure}
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-7-1.pdf}
\caption{Boxplots of scoring distribution for every qualification rank}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-9-1.pdf}
\caption{Kendall's rank correlations - 2018 World Championship, Women's
Qualification}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-11-1.pdf}
\caption{Original rankings and rankings after each rank is dropped}
\end{figure}

>>>>

% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{textcomp}
\usepackage{float}
\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

\if0\blind
{
  \title{\bf An Examination of Sport Climbing Competition Format and
Scoring System}

@@

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf An Examination of Sport Climbing Competition Format and
Scoring System}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
The purpose of this paper is to investigate sport climbing, one of the
sports making its debut on the Olympics stage at Tokyo 2020. In
particular, we take a closer look at the controversial competition
format and scoring system of sport climbing.

\ldots{} Simulation. Data Analysis: drop and re-rank, correlations
\end{abstract}

\noindent%
{\it Keywords:} sport climbing, scoring system, 2020 Summer Olympics
\vfill

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

In the Olympic sport climbing, there are 20 competitors at the start (in
both men's and women's). All 20 competitors compete in each of the three
events in the qualification round, and their performances in each event
are ranked from 1 to 20. A competitors final score is then computed as
the product of their ranks in the three events and the lower product is
better. Specifically, \begin{equation}
Score_i = R^S_i\times R^B_i\times R^L_i,
\end{equation} where \(R^S_i\), \(R^B_i\), and \(R^L_i\) are the ranks
of the \(i\)-th competitor in speed climbing, bouldering, and lead
climbing, respectively.

@@

We collected data on major climbing competitions from 2018 to 2020,
including the 2020 Continental Championships of Europe, Africa, Oceania,
Pan-America; 2019 and 2018 World Championships; 2018 Asian Games; and
2018 Youth Olympics. Data were obtained from various sources, including
the event websites, Wikipedia, and the International Federation of Sport
Climbing (IFSC). The main attributes of our datasets are the name and
nationality of the climbers; bib number (for some competitions); the
finishing place of climbers in speed climbing, bouldering, and lead
climbing; the total score (which equals the product of event ranks); and
the final rank. We utilize this data to compute the correlations between
the event ranks and final table position, as well as to look at how
often the final orderings change if one athlete is dropped and the ranks
for each discipline are re-computed.

EDA

Corr

Drop 1

@@

We conducted a simulation study to examine the rankings and scoring for
climbers in both qualification and final rounds. For each round, we
performed 10000 simulations, and this was accomplished by randomly
assigning the ranks of each event to every participant, with the
assumption that the ranks are uniformly distributed. After the
completion of the simulations, we calculated the total scores for every
simulated round, as well as the final standings for the climbing
athletes. Table and Figure are numerical and visual summaries of final
scores obtaining from our simulated data for the qualification round.

\begin{table}[ht]
\centering
\caption{Descriptive statistics of simulated qualification scores} 
\begin{tabular}{rrrrrrrr}
  \hline
min & Q1 & median & Q3 & max & mean & sd & n \\ 
  \hline
1.00 & 240.00 & 684.00 & 1638.00 & 8000.00 & 1158.84 & 1273.48 & 200000 \\ 
   \hline
\end{tabular}
\end{table}

\begin{figure}[H]
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Histogram of qualification scores}
\end{figure}

The simulation results allow to answer questions about various topics,
including the distributions of scores for qualifying and final rounds,
and the probabilities of advancing to the finals or winning a medal,
given certain conditions. For the qualification round, a climber is
almost guaranteed to make the final round if they win the first event
(with a 99.51\% chance of advancing) or if they win at least one of the
three climbing concentrations (99.48\%). On the other hand, finishing
last in the first event or in any event would certainly hurt an
athlete's chance of finishing in the top 8, as the probabilities of a
climber advancing given they finish last in the first and in any event
are 0.1830 and 0.1885, respectively. In addition, the average score for
qualification positions 1 to 8 are displayed in Table 1. We notice that
on average, the minimum score that one should aim for in order to move
on to the final round is 435 (for 8th rank).

@@

\begin{figure}[H]
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-8-1.pdf}
\caption{Boxplots of scoring distribution for every qualification rank}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-10-1.pdf}
\caption{Kendall's rank correlations - 2018 World Championship, Women's
Qualification}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-12-1.pdf}
\caption{This figure shows the rankings of the Men's Sport Climbing
Final at the 2018 Youth Olympics. Each panel represents the dropped
rank, with 0 being the original final standings. The cases where a
change in ordering happened are highlighted.}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-13-1.pdf}
\caption{This figure shows the final rankings of 2018 Youth Olympics
Women's Final. Each panel represents the dropped rank, with 0 being the
original final standings. The cases where a change in ordering happened
are highlighted.}
\end{figure}
